Measurement Clusterization in Bayesian D-optimal Designs in Infinite Dimensions

### 2024-06-23 latex/intro.tex
changing paper and makin slides

Inverse problems are ubiquitous in science and engineering, and we
will now give several illustrative examples. In electric impedance
tomography (EIT) we seek to infer the impedance of the inside of a
human body from observations of electrical current impedence measured
at specific electrode locations on the skin, leveraging known
equations of electric current flow
\cite{horesh2010impedance}. Magnetic Resonance Imaging (MRI) also
entails solving an inverse problem, where radio-frequency pulses are
sent through the human body in the presence of a strong magnetic field
and the response of our bodies' contents to such perutrbations is
measured, allowing a radiologist to view the internals of a body in a
noninvasive manner \cite{horesh2008mri}. Yet another example of an
inverse problem arises in oil and gas exploration, where acoustic
waves are sent through "boreholes" --- deep cylindrical wells drilled
into the ground --- and data of travel and return times of said waves
are recorded. Since travel and return time is influenced by the
properties of the subsurface materials, such as density, elasticity,
and the presence of fluids or voids, we can combine travel data with a
geophisical models of contents of the earth's crust to facilitate
reconstructing and inferring the structure of the subsurface in a
process called \emph{borehole tomography}
\cite{horesh2008borehole}. Inverse problems also arise in many other
areas of seismology and geology \cite{rabinowitz1990, steinberg1995}
and medical imaging \cite{tarantola}. Importantly, note that in the
above examples our goal is to infer some function $f \in C(\Omega)$,
where \(\Omega \subseteq \mathbb{R}^d, d=1,2,3\) is a spatial domain
of interest.

@@

We will now introduce a toy model which we will use to define a toy
inverse problem: the heat equation in one dimension with a homogeneous
Dirichlet boundary condition. The equations are presented below for
completeness, but readers less familiar with partial differential
equations should think of the heat equation as follows: before us
there is a perfectly insulated metal rod. At each point on the rod the
temperature is different, and we call this temperature distribution
the \emph{initial condition}. We wait for a short time and let the
heat dissipate a bit inside the rod. The final heat distribution in
the rod is dictated by the \emph{1D heat equation}:
\begin{subequations}\label{eq:heat equation}
  \begin{alignat}{2}
    u_t &= \Delta u &&\qquad \text{in } [0,1] \times [0,\infty),\\ u
      &= 0 &&\qquad \text{on } \{0, 1\} \times [0,\infty),\\ u &= u_0
        &&\qquad \text{on }[0,1] \times \{0\}.
  \end{alignat}
\end{subequations}

In the inverse problem, our goal is to infer the initial condition
(i.e.~initial temperature distribution inside the rod). We are allowed
to take some number of temperature readings on the rod once our
waiting time has passed. The inverse problem of the 1D heat equation
is difficult ("ill-posed") since heat spreads in a diffusive manner
and any roughness in the initial condition is quickly smoothed, see
Supplementary movies S1 and S2. We will give a fully Bayesian
formulation of this inverse problem in Section \ref{section:prelims}.

In all of the examples of inverse problems we have seen, as well as in
many other applications, measurements should be chosen to most benefit
the user. Hence, the user may consider some utility, called a
\emph{design criterion} and take measurements that maximize this
utility. Two of the most widely recognized and extensively studied
design criteria are A- and D-optimality. We focus on the Bayesian
D-optimality criterion, which carries a simple and intuitive meaning:
measurements chosen according to the D-optimality criterion maximize
the expected Kullback-Liebler divergence from posterior to prior
\cite{Chaloner1995, AlexanderianGloorGhattas14}, see also precise
definition in Section \ref{subsec:D optimal design}. We refer to a set
of measuremnts that maximizes the Bayesian D-optimality criterion a
\emph{D-optimal design}.

Surprisingly, A- and D-optimal designs for inverse problems have been
observed to yield remarkably similar measurements in certain cases
\cite{fedorov1996, nyberg2012, fedorov1997, Ucinski05,
  neitzel2019sparse}. This phenomenon is illustrated for our toy
inverse problem in Fig.~\ref{fig:clusterization illustration}, where
D-optimal measurement locations are shown for different numbers of
measurements. Notably, for six measurements, a D-optimal design yields
two sets of measurements that are almost indistinguishable from one
another. We refer to this intriguing phenomenon as \emph{measurement
clusterization} \cite{Ucinski05}. For our purposes, it suffices for
measurements to be visibly close for clusterization to hold. We refer
to a design that exhibits measurement clusterization as a
\emph{clustered design},

%% first: blue, second: red, third: green, fourth: black,
%% fifth:magenta
\begin{figure}
    \centering
    \includegraphics[height=0.5\textwidth]{example.pdf}
    \caption{Measurement clusterization for D-optimal designs for the
      inverse problem of the heat equation. Measurement locations were
      chosen according to the Bayesian D-optimality criterion of
      Theorem \ref{thm:d optimality}. Measurement locations are
      plotted over the computational domain \(\Omega = [0, 1]\)
      (x-axis), for varying numbers of measurements (y-axis). The
      colored numbers are measurement indices, plotted for visual
      clarity. Measurement clusterization already occurs for three
      measurements: the second measurement (red) is overlaid on the
      third (green). For five measurements, first (blue) and second
      (red) measurements are clustered, as well as the fourth (black)
      and the fifth (magenta).}
  \label{fig:clusterization illustration}
\end{figure}

@@

In spatial problems involving choice of measurements within a domain
\(\Omega \subseteq \mathbb{R}^d, d=1,2,3\), many researchers
circumvent the problem of measurement clusterization by choosing
measurements from a coarse grid in \(\Omega\) \cite{koval2020,
  alexanderian2021, attia2022, alexanderian2014, alexanderian2016,
  alexanderian2018efficient, brunton2016}. This approach incurs a
significant computational cost as it requires solving a difficult
combinatorial optimization problem for measurement locations over a
finite set. The combinatorial optimization problem is usually relaxed
by first assigning optimal measurement weights in \(\mathbb{R}_+\) to
the potential measurement locations. Some researchers incorporate a
sparsifying \(\ell_1\) penalty term into the design criterion, which
is subsequently thresholded to achieve the desired binary design over
the coarse grid \cite{horesh2008borehole}. Others progressively relax
the \(\ell_1\) penalty to an \(\ell_0\) penalty via a continuation
method \cite{alexanderian2016, alexanderian2014}. Others cast the
problem of finding optimal measurement weights as a stochastic
optimization problem \cite{attia2022stochastic}. All of the
aforementioned methods may indeed find a binary optimal design
restricted to a given coarse grid. However, none addresses one
fundamental issue: the restriction of measurement locations to a
coarse grid in \(\Omega\) fundamentally changes the optimal design
problem and thus results in a sub-optimal design.

@@

\subsection{Contribution}
The primary objective of this study is to provide a comprehensive
understanding of measurement clusterization by addressing the
aforementioned questions. Our focus centers around investigating the
Bayesian D-optimality criterion, which involves maximizing the
expected Kullback-Leibler divergence between the posterior and prior
measures \cite{CoverThomas91, Chaloner1995}. We conduct an analysis of
Bayesian D-optimal designs within the context of linear inverse
problems over Hilbert spaces. We propose a novel relaxed model for
D-optimality that maintains analytical tractability and enables the
identification of D-optimal designs using Lagrange multipliers. This
analytical framework facilitates the exploration of the questions
posed at the end of the previous paragraph:

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{eigenvectors.pdf}
    \caption{D-optimal measurement locations ($m=4$ measurements) and
      weighted eigenvectors for finding the initial condition of
      the 1D heat equation. Measurement locations and weighted
      eigenvectors are plotted over the computational domain $\Omega =
      [0, 1]$ (x-axis). Measurement clusterization occurs
      approximately at $0.31$ and $0.69$. These two locations are a
      compromise between zeros of eigenvectors a D-optimal design aims
      to ignore (third and up) and staying far from zeros of the
      eigenvectors a D-optimal design aims to measure (first and
      second). Allocating $m=4$ measurements into two locations
      results in clusterization, according to the pigeonhole
      principle.}
  \label{fig:why}
\end{figure}

\begin{enumerate}

\item \label{q:generic} \textbf{Is measurement clusterization a
  generic phenomenon?}
  %\subsection{An answer for Question \ref{q:generic}: Genericity of measurement clusterization}
  %% Computer implementation of Lemma \ref{thm:char} for the inverse
  %% problem outlined in Section \ref{section:how} also generates \(\obs\)
  %% as a solution to the D-optimal design problem.  Furthermore,
  Randomized numerical simulations of our model give rise to D-optimal
  designs that exhibit clusterization more than 95\% of the time (see
  code in supplementary material). Given our model's genericity, we
  expect measurement clusterization to be a generic phenomenon.

\item \label{q:mitigate} \textbf{Why does imposing correlations
  between observations alleviate measurement clusterization?} In
  Section \ref{section:non vanishing}, we rigorously demonstrate the
  role of model error in mitigating clusterization, thereby
  corroborating earlier observations made by various researchers.

\item \label{q:why} \textbf{Why does measurement clusterization
  occur?} In Section \ref{section:why}, we present a compelling
  explanation for the optimality of clustered designs in the absence
  of model error. Our analysis reveals that, in our model, a D-optimal
  design focuses on a select set of prior eigenvectors, specifically
  those with the largest eigenvalues in the prior covariance
  spectrum. In practical scenarios, the number of locations where (a)
  the relevant prior eigenvectors are significantly large, and (b)
  other eigenvectors are close to zero, is limited. Consequently, the
  clusterization of measurements arises as a natural consequence of
  the pigeonhole principle, as there are more measurements available
  than there are locations satisfying conditions (a) and (b). See
  Fig~\ref{fig:why}.

@@

The first insight our study gives is that a D-optimal design aims to
uniformly reduce uncertainties not over space (i.e.~$\Omega$), but
rather over \emph{eigenvectors}. This insight is a part of Theorem
\ref{thm:char}, see Section \ref{section:vanishing} for precise
statement and proof, and see Fig.~\ref{cluste} for an illustration.

In Theorem \ref{thm:char}, we also show that D-optimal designs are
best understood in the sapce of \emph{observations}. Again, a precise
statement could be found in Section \ref{section:vanishing}.  This
result is in accordance with previous work by \cite{koval2020}, who
showed that A-optimal designs are best constructed in the space of
observations.

@@

\subsection{Limitations}\label{subsec:limitations}
The main limitation of this study is that our generic model does not
correspond to any specific real-life problem. It is generic enough to
be analytically tractable, but one may argue our model is too far
removed from any real application. To these claims I would answer that
scientists have a long history of studying models that are bare-bones
simplifications of real systems, e.g. the Ising model
\cite{cipra1987}, the Lorenz system \cite{brin}, the Lotka-Volterra
equations \cite{logan2006}, the Carnot engine \cite{kardar2007}, and
many others.

>>>>

Inverse problems are ubiquitous in science and engineering, and we
will now give several illustrative examples. In electric impedance
tomography (EIT) we seek to infer the impedance of the inside of a
human body from observations of electrical current impedence measured
at specific electrode locations on the skin, leveraging known
equations of electric current flow
\cite{horesh2010impedance}. Magnetic Resonance Imaging (MRI) also
entails solving an inverse problem, where radio-frequency pulses are
sent through the human body in the presence of a strong magnetic field
and the response of our bodies' contents to such perutrbations is
measured, allowing a radiologist to view the internals of a body in a
noninvasive manner \cite{horesh2008mri}. Yet another example of an
inverse problem arises in oil and gas exploration, where acoustic
waves are sent through "boreholes" --- deep cylindrical wells drilled
into the ground --- and data of travel and return times of said waves
are recorded. Since travel and return time is influenced by the
properties of the subsurface materials, such as density, elasticity,
and the presence of fluids or voids, we can combine travel data with a
geophisical models of contents of the earth's crust to facilitate
reconstructing and inferring the structure of the subsurface in a
process called \emph{borehole tomography}
\cite{horesh2008borehole}. Inverse problems also arise in many other
areas of seismology and geology \cite{rabinowitz1990, steinberg1995}
and medical imaging \cite{tarantola}. While in many inverse problems
the goal is to infer some parameter, in all of the above examples and
in the formulation we consider here, the goal is to infer some
\emph{function} over $ \Omega$, where \(\Omega \subseteq \mathbb{R}^d,
d=1,2,3\) is the spatial domain of interest.

@@

For concreteness, let us consider a toy model of an inverse problem:
the \emph{heat equation} in one dimension (1D heat equation
henceforth, see Section \ref{subsec:1d heat} for more details). The
heat equation is a partial differential equation which is in fact a
set of three equations: eqs.~\eqref{eq:heat1}, \eqref{eq:heat2}, and
\eqref{eq:heat3}. Readers less familiar with partial differential
equations should think of the heat equation as follows: before us
there is a perfectly insulated metal rod. At each point on the rod the
temperature is different and unknown, and we call this temperature
distribution the \emph{initial condition}. We wait for a short time
and let the heat dissipate a bit inside the rod. The final heat
distribution in the rod is dictated by the following set of equations
which defines the 1D heat equation:
\begin{subequations}%\label{eq:heat equation}
  \begin{alignat}{2}
    u_t &= \alpha \Delta u &&\qquad \text{in } [0,1] \times [0,\infty), \label{eq:heat1}\\
    u &= 0 &&\qquad \text{on } \{0, 1\} \times [0,\infty), \label{eq:heat2}\\
    u &= u_0 &&\qquad \text{on }[0,1] \times \{0\}. \label{eq:heat3}\\
  \end{alignat}
\end{subequations}

As time passes, heat disspates across the rod as hotter regions,
e.g.~local maximas of the heat distribution, become cooler. This
behavior is captured by eq.~\eqref{eq:heat1}: the Laplacian at local
maxima is negative, so $u_t = \Delta u$ implies $u$ decreases at its
maximas, and the reverse happens at local minima. Eq.~\eqref{eq:heat2}
describes the \emph{boundary condition}, i.e.~the interaction of the
heat inside the rod with the boundary of the domain $\Omega =
[0,1]$. Here we consider a \emph{homogeneous Dirichlet boundary
condition} which is best understood as an absorbing boundary: heat
that interacts with the boundary immediately
disappears. Eq.~\eqref{eq:heat3} describes the \emph{initial
condition} of the heat, i.e.~the heat distribution at time $t=0$.

So far, we described the \emph{forward} problem: how heat evolves in
time. In the inverse problem, our goal is to infer the initial
condition $u_0 = u(\cdot, 0)$ from observations of the final condition
$u(\cdot, T)$. We are allowed to take some number of temperature
measurements on the rod once our waiting time has passed. The inverse
problem of the 1D heat equation is difficult ("ill-posed") since heat
spreads in a diffusive manner and any roughness in the initial
condition is quickly smoothed, see Supplementary movies S1 and S2. We
will give a fully Bayesian formulation of a generic linear inverse
problem in Section \ref{section:prelims}.

In all of the examples of inverse problems we have seen, as well as in
many other applications, measurements should be chosen to most benefit
the user. Hence, the user may consider some utility, called a
\emph{design criterion} and take measurements that maximize this
utility. Two of the most widely recognized and extensively studied
design criteria are A- and D-optimality. We focus on the Bayesian
D-optimality criterion, which carries a simple and intuitive meaning:
measurements chosen according to the D-optimality criterion maximize
the expected Kullback-Liebler divergence from posterior to prior
\cite{Chaloner1995, AlexanderianGloorGhattas14}, see Section
\ref{subsec:D optimal design} for a precise definition. We refer to a
set of measuremnts that maximizes the Bayesian D-optimality criterion
a \emph{D-optimal design}.

Surprisingly, A- and D-optimal designs for inverse problems have been
observed to yield remarkably similar measurements in certain cases
\cite{fedorov1996, nyberg2012, fedorov1997, Ucinski05,
  neitzel2019sparse}. This phenomenon is illustrated for our toy
inverse problem in Fig.~\ref{fig:clusterization illustration}, where
D-optimal measurement locations are shown for different numbers of
measurements. Notably, for six measurements, a D-optimal design yields
two sets of measurements that are almost indistinguishable from one
another. We refer to this intriguing phenomenon as \emph{measurement
clusterization} \cite{Ucinski05}. We consider a design to be
\emph{clustered} when two or more of its constituent measurements are
identical.

%% first: blue, second: red, third: green, fourth: black,
%% fifth:magenta
\begin{figure}
    \centering
    \includegraphics[height=0.5\textwidth]{dst_modelError0.pdf}
    \caption{Measurement clusterization for D-optimal designs for the
      inverse problem of the heat equation. Measurement locations were
      chosen according to the Bayesian D-optimality criterion of
      Theorem \ref{thm:d optimality}. Measurement locations are
      plotted over the computational domain \(\Omega = [0, 1]\)
      (x-axis), for varying numbers of measurements (y-axis). The
      colored numbers are measurement indices, plotted for visual
      clarity. Measurement clusterization already occurs for three
      measurements: the second measurement (red) is overlaid on the
      third (green). For five measurements, first (blue) and second
      (red) measurements are clustered, as well as the fourth (black)
      and the fifth (magenta).}
  \label{fig:clusterization illustration}
\end{figure}

@@

In spatial problems involving choice of measurements within a domain
\(\Omega \subseteq \mathbb{R}^d, d=1,2,3\), many researchers
circumvent the problem of measurement clusterization by choosing
measurements from a coarse grid in \(\Omega\) \cite{koval2020,
  alexanderian2021, attia2022, alexanderian2014, alexanderian2016,
  alexanderian2018efficient, brunton2016}. This approach incurs a
significant computational cost as it requires solving a difficult
combinatorial optimization problem for measurement locations over a
discrete set. The combinatorial optimization problem is usually
relaxed by first assigning optimal measurement weights in
\(\mathbb{R}_+\) to the potential measurement locations. Some
researchers incorporate a sparsifying \(\ell_1\) penalty term into the
design criterion, which is subsequently thresholded to achieve the
desired binary design over the coarse grid
\cite{horesh2008borehole}. Others progressively relax the \(\ell_1\)
penalty to an \(\ell_0\) penalty via a continuation method
\cite{alexanderian2016, alexanderian2014}. Others cast the problem of
finding optimal measurement weights as a stochastic optimization
problem \cite{attia2022stochastic}. All of the aforementioned methods
may indeed find a binary optimal design restricted to a given coarse
grid. However, none addresses one fundamental issue: the restriction
of measurement locations to a coarse grid in \(\Omega\) fundamentally
changes the optimal design problem and thus results in a sub-optimal
design.

@@

\subsection{Contribution}
The primary objective of this study is to provide a comprehensive
understanding of measurement clusterization by addressing the
aforementioned questions. Our focus centers around investigating the
Bayesian D-optimality criterion. We conduct an analysis of Bayesian
D-optimal designs within the context of linear inverse problems over
Hilbert spaces. We propose a novel relaxed model for D-optimality that
maintains analytical tractability and enables the identification of
D-optimal designs using Lagrange multipliers. This analytical
framework facilitates the exploration of the questions posed at the
end of the previous paragraph:

\begin{enumerate}
\item \label{q:generic} \textbf{Is measurement clusterization a
  generic phenomenon?}
  %\subsection{An answer for Question \ref{q:generic}: Genericity of measurement clusterization}
  %% Computer implementation of Lemma \ref{thm:char} for the inverse
  %% problem outlined in Section \ref{section:how} also generates \(\obs\)
  %% as a solution to the D-optimal design problem.  Furthermore,
  Randomized numerical simulations of our model give rise to D-optimal
  designs that exhibit clusterization more than 95\% of the time (see
  Section \ref{subsec:lemma sims} for details). Given the genericity
  of our model, we expect measurement clusterization to be a generic
  and ubiquitous phenomenon.

\item \label{q:mitigate} \textbf{Why does imposing correlations
  between observations alleviate measurement clusterization?} In
  Section \ref{section:non vanishing}, we rigorously demonstrate the
  role of model error in mitigating clusterization, thereby
  corroborating earlier observations made by various researchers. We
  also show that adding to the inverse problem of the 1D heat
  equationan error term corresponding to correlations between
  measurements mitigates clusterization, see Section \ref{subsec:corr
    errors sims}.

\item \label{q:why} \textbf{Why does measurement clusterization
  occur?} Answering this question is the main contribution of this
  study, and it is arrived at in Section \ref{section:why}. There, we
  present a compelling explanation for the optimality of clustered
  designs in the absence of model error. Our analysis reveals that a
  D-optimal design focuses on a select set of prior eigenvectors,
  specifically those with the largest eigenvalues in the prior
  covariance spectrum. In practical scenarios, the number of locations
  where (a) the relevant prior eigenvectors are significantly large,
  and (b) other eigenvectors are close to zero, is
  limited. Consequently, the clusterization of measurements arises as
  a natural consequence of the pigeonhole principle, as there are more
  measurements available than there are locations satisfying
  conditions (a) and (b). See Fig~\ref{fig:why}.

@@

The implication of the answers we give to Questions \ref{q:generic},
\ref{q:mitigate}, and \ref{q:why} is two-fold:

The implication of our answer to Question \ref{q:generic} is that
encountering clusterization should be expected in many different
inverse problems across many different scientific fields and
disciplines. Thus, researchers that encounter clusterization should
not be surprised or wary. Then, in our answer to Question \ref{q:why},
we explain what is the source of clusterization. As it appears, the
source is universal: a D-optimal design reduces uncertainty for a
select set of prior covariance eigenvectors --- those with the most
prior uncertainty, i.e.~those the practitioner cares about the most!
Consequently, practitioners should not try to avoid measurement
clusterization. Rather, practitioners should should take repeated
measurements (e.g.~in MRI and borehole tomography), increase apparatus
sensitivity (e.g.~in EIT), or take consecutive measurements (e.g.~in
the 1D heat equation). Overall, we show that measurement
clusterization is a natural and (almost) inevitable part of Bayesian
D-optimal designs.

\subsubsection{Further Contributions}
In Theorem \ref{thm:char}, we also show that D-optimal designs are
best understood in the sapce of \emph{observations}. Again, a precise
statement could be found in Section \ref{section:vanishing}.  This
result is in accordance with previous work by \cite{koval2020}, who
showed that A-optimal designs are best constructed in the space of
observations.

@@

\subsection{Limitations}\label{subsec:limitations}
The main limitation of this study is that our generic model does not
correspond to any specific real-life problem. Specifically, in its
current form, our model does not allow point evaluations, see the
Discussion for a deeper analysis of this subject. Our model is generic
enough to be analytically tractable, but one may argue our model is
too far removed from any real application. To these claims I would
answer that scientists have a long history of studying models that are
bare-bones simplifications of real systems, e.g. the Ising model
\cite{cipra1987}, the Lorenz system \cite{brin}, the Lotka-Volterra
equations \cite{logan2006}, the Carnot engine \cite{kardar2007}, and
many others.

### 2024-06-23 latex/numerix.tex
changing paper and makin slides

\subsection{Heat Equation}
Our goal is to infer the initial condition of
the one-dimensional heat equation with a homogeneous Dirichlet
boundary condition:
\begin{subequations}\label{eq:heat equation}
  \begin{alignat}{2}
    u_t &= \Delta u &&\qquad \text{in } [0,1] \times [0,\infty),\\
      u &= 0 &&\qquad \text{on } \{0, 1\} \times [0,\infty),\\
        u &= u_0 &&\qquad \text{on }[0,1] \times \{0\}.
  \end{alignat}
\end{subequations}

Model the prior for the initial condition as $u_0 \sim
\normal(0,\prcov)$, for $\prcov = (-\Delta)^{-1}$ with a homogeneous
Dirichlet boundary condition. Let the final time $T = 0.03$ and denote
$\fwd$ the forward operator, so that $u(\cdot, T) = \fwd u_0$. Let
$\obs$ a measurement operator, so that $u(\x_j,T) = (\obs u)_j,
j=1,\dots,m$, where $m$ is the number of allowed measurements. Assume
iid centered Gaussian observation error, so observations are given by
$u(\x_j,T) + \eps(\x_j)$ with $\eps(\x_j) \sim \normal(0, \sigma^2)$
iid, $\sigma = 0.05$. Measurement locations $\x_j,j=1,\dots,m$ are
chosen according to the D-optimality criterion
\cite{AlexanderianGloorGhattas14}.

Implementation of this inverse problem can be found in this 
\href{https://github.com/yairdaon/OED}{repository}, including an optimization
routine for finding D-optimal designs. Unfortunately, I was not able
to install PyOED \cite{attia2023pyoed}, nor was I able to utilize
hIPPYlib \cite{VillaPetraGhattas16, VillaPetraGhattas18,
  VillaPetraGhattas21}, hence the inverse problem is implemented from
scratch.

\subsection{Simulating Lemma ... }
In theorem 17 we explicitly construct a D-optimal design in a Hilbert
space where the base vectors diagonalize $\fwd \prcov\fwd^*$. We
explicitly implement the algorithm outlined in Lemma

\subsection{Correlated errors}
We simulate the 1D heat eq

>>>>

\subsection{Simulating Lemma \ref{lemma:free}}\label{subsec:lemma sims}
Lemma \ref{lemma:free} is used to construct D-optimal designs as a
part of the proof of Theorem \ref{thm:char}. We implement the
construction in Lemma \ref{lemma:free} (see module \texttt{zeros.py}
in the accompanying
\href{https://github.com/yairdaon/OED}{repository}). We iterate over
the number of measurements $m \in \{4,\dots, 15\}$, and for every $m$
we then iterate over $k:=\rank \obs^*\obs \in \{2,\dots, m-1\}$. For
each pair $m,k$ we run $N=50000$ simulations according to the
following steps:
\begin{enumerate}
\item Generate random diagonal $D\in \mathbb{R}^{k\times k}$ and
  normalize it so that $\ttr D = m$.
\item Conjugate $D$ by a random orthogonal matrix to form a positive
  semi-definite $M := UDU^t \in \mathbb{R}^{k\times k}$.
\item Apply the construction of Lemma \ref{lemma:free} to calculate
  $A$ such that $AA^t = M$, where $A$ has unit norm columns.
\item We call $A$ "clustered" if $A$ has two or more identical columns
  (up to some numerical precision threshold).
\end{enumerate}
We then calculate the fraction of clustered designs of the $N=5000$
simulations we ran, for each pair $m,k$. Full results in the
\texttt{simulations.csv} file within the accompanying
\href{https://github.com/yairdaon/OED}{repository}), below we give a
brief description of the results.

Clusteriztion occured at high rates ($>95\%$) whenever $m-k > 1$; see
Fig.~\ref{fig:sim AAt}. Hence, in these simulations, clusterization is
a generic property. However, when $m-k \leq 2$, clusterization does
not occur. We do not why and further investigation into this
phenomenon is out of scope for the current study.

\begin{figure}
    \centering
    \includegraphics[height=0.5\textwidth]{simulations.png}
    \caption{Fraction of clustered $A$ for $AA^t = M$ and $M$
      generated randomly (see text and repository for details on
      generating $M$). It is evident that when $m-k >3$ clusterization
      is ubiquitous, whereas for lower $m-k$ clusterization does not
      occur.}
  \label{fig:sim AAt}
\end{figure}

\subsection{Correlated errors}\label{subsec:corr errors sims}
In order to verify the results of Section \ref{section:non vanishing},
we run simulations of the inverse problem of the 1D heat equation
\ref{subsec:1d heat} with nonvanishing model error. We take a model
correlation of \(\modcov = \prcov^2 \). Indeed, adding model
correlation pushes measurements apart as can be seen in
Fig.~\ref{fig:corr errors}. Code generating the image below can be
located at module \texttt{clusterization.py} in the accompanying
\href{https://github.com/yairdaon/OED}{repository}).

\begin{figure}
    \centering
    \includegraphics[height=0.5\textwidth]{dst_modelError4.pdf}
    \caption{Model correlation mitigates clusterization. We add a
      model correlation term to the error terms in the 1D heat
      equation inverse problem. Lo and behold, measurements are not
      close anymore and are pushed away thanks to the model error
      term.}
  \label{fig:corr errors}
\end{figure}

### 2024-06-23 latex/prelims.tex
changing paper and makin slides

\section{Preliminaries and Notation}\label{section:prelim}

In this section we present the setup that will be used throughout this
article. The theoretical foundations for inverse problems over
function spaces can be found elsewhere \cite{Stuart10} and will not be
reviewed here.

\subsection{Bayesian Linear Inverse Problems}\label{subsec:abstract OED}
Let \(\hilp\) and \(\hilo\) separable Hilbert spaces (the subscripts p
and o are for "parameter" and "observation", respectively), and let
\(\fwd: \hilp \to \hilo\) the \emph{forward operator}. The forward
operator \(\fwd\) is assumed linear with fast decaying singular values
(the heat operator of the 1D heat equation introduced in Section
\ref{section:intro} is a prime example). Take a Gaussian prior
\(\param \sim \pr = \normal(\prmean ,\prcov)\) with some appropriate
covariance operator \(\prcov\) on \(\hilp\) \cite{Stuart10}. Note that
\(\fwd \prcov \fwd^*\) is the prior covariance in \(\hilo\)
\cite{Stuart10}, and as such, \(\fwd \prcov \fwd^*\) is assumed
invertible --- an assumption which will be used below (if \(\fwd\) has
a nontrivial kernel we utilize Occam's Razor and ignore said kernel
altogether). Measurements are taken via a \emph{measurement operator}
\(\obs\). It is common for the measurement and forward operators to be
merged \(\tmp := \obs \fwd\) \cite{AlexanderianGloorGhattas14}, but
the analysis carried out in the following sections requires that
\(\fwd\) and \(\obs\) are explicitly separated as in
\cite{attia2022stochastic, cvetkovic2023choosing}.  \(\obs \in (
\hilo^* )^m\), where \(m\) is the number of measurements
taken. Entries \(\meas_j, j=1,\dots,m\) of the observation operator
\(\obs\) are called \emph{measurements}:

@@

Finally, it is useful to record that the posterior measure \(\post\)
is Gaussian in this setting and its covariance operator does not
depend on data \(\data\) \cite{Stuart10}:
\begin{align}\label{eq:postcov}
  \postcov = (\prcov^{-1} + \fwd^* \obs^* \Sigma^{-1} \obs \fwd
  )^{-1}.
\end{align}

\subsection{Bayesian D-Optimal Designs in Infinite Dimensions}\label{subsec:D optimal design} 
A Bayesian D-optimal design maximizes expected Kullback-Leibler (KL)
divergence between posterior \(\post\) and prior measures \(\pr\). It is
useful to first recall the definition of KL divergence \cite{CoverThomas91}:
\begin{equation*}
  D_{KL}(\post||\pr) = \int \log \frac{\der \post}{\der \pr}(\param) \der \post(\param).
\end{equation*}

>>>>

\section{Preliminaries and Notation}\label{section:prelim}
Here we present the basics of Bayesian inverse problems over hilbert
spaces. Since we are ultimately interested in inferring a function over some
domain, we will keep in mind that these Hilbert spaces should really
be thought of as function spaces. The theoretical foundations for
inverse problems over function spaces can be found in \cite{Stuart10}
and will be only briefly reviewed here.

\subsection{Forward Problems}\label{subsec:abstract OED}
In this section we give definitions and notations for forward probles,
which are an essential part of the inverse problems we discuss
later. Consider a "parameter space" \(\hilp\) and an "observation
space" \(\hilo\) --- both separable Hilbert spaces (the subscripts p
and o are for "parameter" and "observation", respectively). The
parameter space includes the quantity we seek to infer; in the inverse
problem of the heat equation, the parameter space $\hilp$ is where the
initial condition lives. The observation space $\hilo$, on the other
hand, is the space from which we take measurements; in the example of
the 1D heat equation, $u(\cdot, T) \in \hilo$.

The connection between parameter and observation spaces is made by the
\emph{forward operator} \(\fwd: \hilp \to \hilo\). We assume the
forward operator \(\fwd\) is linear. In the inverse problem of the 1D
heat, the forward operator is determined by the time evolution of the
1D heat equation \eqref{eq:heat1}, \eqref{eq:heat2} and
\eqref{eq:heat3}.

Measurements are taken via a \emph{measurement operator} \(\obs\). It
is common for the measurement and forward operators to be merged
\(\tmp := \obs \fwd\) \cite{AlexanderianGloorGhattas14}, but the
analysis carried out in the following sections requires that \(\fwd\)
and \(\obs\) are explicitly separated (as in
\cite{attia2022stochastic, cvetkovic2023choosing}). Note that \(\obs
\in ( \hilo^* )^m\), where \(m\) is the number of measurements
taken\footnote{The alert reader will likely ask how do we reconcile
point measurements $\delta_x$ as suggested by the formulation of the
1D heat equation. We don't. This entire study could be formulated for
Banach spaces where point measurements are allowed. We avoid this
technical difficulty because: (1) it does not add much to the study,
(2) it is beyond the technical abilities of the author}. Entries
\(\meas_j, j=1,\dots,m\) of the measurement operator \(\obs\) are
called \emph{measurements}:

@@

\subsection{Bayesian Linear Inverse Problems}\label{subsec:bayesian inverse problems}
In the previous section, we saw how a parameter $u\in \hilp$ is
transported to the observation space via the forward operaror $\fwd u
\in \hilo$, how observations are generated via $\obs \fwd u$ and how
observations and noise give rise to data $\data$. It is time to
formulate the process of inferring the parameter as a Bayesian inverse
problem. The first ingredient we need is a prior.

We take a Gaussian prior \(\param \sim \pr = \normal(\prmean
,\prcov)\) with some appropriate covariance operator \(\prcov\) on
\(\hilp\) \cite{Stuart10}. For the inverse problem of the 1D heat
equation we choose $\prcov = -\Delta^{-1.1}$, see Section
\ref{subsec:1d heat} for details and a short discussion of
"Laplacian-like" covariance operators. Note that \(\fwd \prcov
\fwd^*\) is the prior covariance in \(\hilo\) \cite{Stuart10}, and as
such, \(\fwd \prcov \fwd^*\) is assumed invertible --- an assumption
which will be used below (if \(\fwd\) has a nontrivial kernel we
utilize Occam's Razor and ignore said kernel altogether).

Since this inverse problem is linear and the prior is Gaussian --- the
posterior measure \(\post\) is Gaussian as well. We do not utilize the
posterior mean in this study and posterior covariance operator
$\postcov$ is given by \cite{Stuart10}:
\begin{align}\label{eq:postcov}
  \postcov = (\prcov^{-1} + \fwd^* \obs^* \Sigma^{-1} \obs \fwd
  )^{-1}.
\end{align}

\subsection{Bayesian D-Optimal Designs}\label{subsec:D optimal design} 
A Bayesian D-optimal design maximizes expected Kullback-Leibler (KL)
divergence between posterior \(\post\) and prior measures \(\pr\). It is
useful to first recall the definition of KL divergence \cite{CoverThomas91}:
\begin{equation*}
  D_{KL}(\post||\pr) = \int \log \frac{\der \post}{\der \pr}(\param) \der \post(\param).
\end{equation*}
